---
title: "Analysis"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE, echo = TRUE,
  fig.align = "center", fig.width = 8, fig.height = 5
)
```

```{r load-data, include=FALSE}
library(tidyverse)
library(lubridate)
library(scales)
library(sf)
library(knitr)
library(kableExtra)
library(viridis)
library(gridExtra)

load("data/istanbul_airbnb.RData")
```

## Key Takeaways

- **Listings are heavily concentrated in tourist cores:** Beyoglu, Sisli, and Fatih together account for over 45% of all Istanbul listings.
- **Price is driven by room type and location:** Entire-home listings cost 2-3x more than private rooms; central districts command a clear premium.
- **COVID-19 caused a near-total booking collapse:** Monthly review counts fell ~85% in spring 2020, with full recovery by mid-2022.
- **Professional hosts dominate supply:** Hosts with 10+ listings control a disproportionate share of the market.
- **Availability patterns reveal commercialisation:** Multi-listing hosts keep their properties available year-round, unlike casual hosts.

---

## 1. Geographic Distribution

### 1.1 Map of All Listings

```{r map}
listings %>%
  ggplot(aes(x = longitude, y = latitude, colour = room_type)) +
  geom_point(alpha = 0.4, size = 0.8) +
  scale_colour_manual(
    values = c("#e63946", "#457b9d", "#2a9d8f", "#e9c46a"),
    name   = "Room Type"
  ) +
  labs(
    title   = "Geographic Distribution of Istanbul Airbnb Listings",
    x = "Longitude", y = "Latitude",
    caption = "Source: Inside Airbnb - Istanbul"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

Listings cluster around the Bosphorus shoreline, with the densest concentration in Beyoglu and the historic peninsula.

### 1.2 Top 15 Districts by Listing Count

```{r listings-by-district}
listings %>%
  count(neighbourhood, sort = TRUE) %>%
  slice_head(n = 15) %>%
  mutate(neighbourhood = fct_reorder(neighbourhood, n)) %>%
  ggplot(aes(x = n, y = neighbourhood, fill = n)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_c(option = "C") +
  scale_x_continuous(labels = comma) +
  labs(
    title   = "Top 15 Districts by Listing Count",
    x = "Number of Listings", y = NULL,
    caption = "Source: Inside Airbnb - Istanbul"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

Beyoglu dominates with the highest listing count, consistent with its status as Istanbul's main tourist hub.

### 1.3 Choropleth Map

```{r choropleth}
listing_counts <- listings %>%
  count(neighbourhood, name = "n_listings")

map_data <- neighbourhoods_geo %>%
  left_join(listing_counts, by = c("neighbourhood" = "neighbourhood"))

ggplot(map_data) +
  geom_sf(aes(fill = n_listings), colour = "white", linewidth = 0.3) +
  scale_fill_viridis_c(option = "C", na.value = "grey90",
                       labels = comma, name = "Listings") +
  labs(
    title   = "Listing Density Across Istanbul Districts",
    caption = "Source: Inside Airbnb - Istanbul"
  ) +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

## 2. Price Analysis

### 2.1 Price Distribution

```{r price-dist}
listings %>%
  ggplot(aes(x = price)) +
  geom_histogram(bins = 60, fill = "#457b9d", colour = "white") +
  scale_x_log10(labels = dollar) +
  scale_y_continuous(labels = comma) +
  labs(
    title   = "Distribution of Nightly Prices (log scale)",
    x = "Price per Night (USD)", y = "Count",
    caption = "Source: Inside Airbnb - Istanbul"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

The distribution is right-skewed with a median around $20-50 per night.

### 2.2 Price by Room Type

```{r price-room}
listings %>%
  ggplot(aes(x = room_type, y = price, fill = room_type)) +
  geom_boxplot(outlier.alpha = 0.2, show.legend = FALSE) +
  scale_y_log10(labels = dollar) +
  scale_fill_manual(values = c("#e63946","#457b9d","#2a9d8f","#e9c46a")) +
  labs(
    title   = "Nightly Price by Room Type",
    x = NULL, y = "Price (USD, log scale)",
    caption = "Source: Inside Airbnb - Istanbul"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

Entire homes cost roughly 2x the price of private rooms.

### 2.3 Top 15 Districts by Median Price

```{r price-neighbourhood}
listings %>%
  group_by(neighbourhood) %>%
  summarise(median_price = median(price, na.rm = TRUE), n = n()) %>%
  filter(n >= 30) %>%
  slice_max(median_price, n = 15) %>%
  mutate(neighbourhood = fct_reorder(neighbourhood, median_price)) %>%
  ggplot(aes(x = median_price, y = neighbourhood, fill = median_price)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_c(option = "B") +
  scale_x_continuous(labels = dollar) +
  labs(
    title   = "Top 15 Districts by Median Nightly Price",
    x = "Median Price (USD)", y = NULL,
    caption = "Source: Inside Airbnb - Istanbul (districts with 30+ listings)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

### 2.4 Price vs Minimum Nights

```{r price-min-nights}
listings %>%
  filter(minimum_nights <= 30) %>%
  ggplot(aes(x = factor(minimum_nights), y = price)) +
  geom_boxplot(fill = "#457b9d", outlier.alpha = 0.1,
               show.legend = FALSE) +
  scale_y_log10(labels = dollar) +
  labs(
    title   = "Price Distribution by Minimum Night Requirement",
    subtitle = "Listings requiring longer stays tend to have lower nightly prices",
    x = "Minimum Nights", y = "Price (USD, log scale)",
    caption = "Source: Inside Airbnb - Istanbul"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

## 3. Temporal Analysis & COVID-19

Reviews are a reliable proxy for bookings since a review can only be left after a completed stay.

```{r timeseries}
reviews %>%
  filter(year >= 2015, year <= 2024) %>%
  count(year_month = floor_date(date, "month")) %>%
  ggplot(aes(x = year_month, y = n)) +
  geom_line(colour = "#457b9d", linewidth = 1) +
  geom_area(fill = "#457b9d", alpha = 0.2) +
  annotate("rect",
           xmin = as.Date("2020-03-01"), xmax = as.Date("2021-06-01"),
           ymin = -Inf, ymax = Inf, alpha = 0.15, fill = "#e63946") +
  annotate("text", x = as.Date("2020-09-01"), y = 10000,
           label = "COVID-19", colour = "#e63946",
           fontface = "bold", size = 3.5) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Monthly Review Count (2015-2024)",
    subtitle = "~85% collapse in spring 2020; full recovery by mid-2022",
    x = NULL, y = "Reviews",
    caption = "Source: Inside Airbnb - Istanbul"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

### 3.1 Seasonal Patterns

```{r seasonality}
reviews %>%
  filter(year %in% c(2018, 2019, 2022, 2023)) %>%
  count(year, month_num = month(date)) %>%
  mutate(year = factor(year)) %>%
  ggplot(aes(x = month_num, y = n, colour = year, group = year)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_continuous(labels = comma) +
  scale_colour_viridis_d(option = "D") +
  labs(
    title   = "Seasonal Patterns (Selected Years)",
    subtitle = "July-August peak is consistent across all years",
    x = NULL, y = "Reviews", colour = "Year",
    caption = "Source: Inside Airbnb - Istanbul"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

Istanbul's Airbnb market peaks sharply in summer and troughs in January-February.

---

## 4. Host Profile Analysis

### 4.1 Single vs Multi-Listing Hosts

```{r host-profile}
host_summary <- listings %>%
  group_by(host_id) %>%
  summarise(n_listings = n_distinct(id)) %>%
  mutate(host_type = case_when(
    n_listings == 1  ~ "1 listing",
    n_listings <= 4  ~ "2-4 listings",
    n_listings <= 9  ~ "5-9 listings",
    TRUE             ~ "10+ listings"
  ),
  host_type = factor(host_type,
                     levels = c("1 listing", "2-4 listings",
                                "5-9 listings", "10+ listings")))

p1 <- host_summary %>%
  count(host_type) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = host_type, y = pct, fill = host_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)),
            vjust = -0.4, size = 3) +
  scale_y_continuous(labels = percent, limits = c(0, 0.85)) +
  scale_fill_viridis_d(option = "C") +
  labs(title = "Share of Hosts", x = NULL, y = NULL) +
  theme_minimal(base_size = 10)

p2 <- listings %>%
  left_join(host_summary, by = "host_id") %>%
  count(host_type) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = host_type, y = pct, fill = host_type)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)),
            vjust = -0.4, size = 3) +
  scale_y_continuous(labels = percent, limits = c(0, 0.6)) +
  scale_fill_viridis_d(option = "C") +
  labs(title = "Share of Listings Controlled", x = NULL, y = NULL) +
  theme_minimal(base_size = 10)

grid.arrange(p1, p2, ncol = 2)
```

Single-listing hosts are the majority, but multi-listing operators control a disproportionate share of supply.

### 4.2 Availability by Host Type

```{r availability}
listings %>%
  left_join(host_summary, by = "host_id") %>%
  ggplot(aes(x = host_type, y = availability_365, fill = host_type)) +
  geom_boxplot(outlier.alpha = 0.2, show.legend = FALSE) +
  scale_fill_viridis_d(option = "C") +
  labs(
    title   = "Annual Availability by Host Type",
    subtitle = "Professional hosts keep listings available significantly longer",
    x = NULL, y = "Days Available per Year",
    caption = "Source: Inside Airbnb - Istanbul"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

## 5. Tourism Pressure Index

We combine three normalised components into an equal-weight **Tourism Pressure Index (TPI)**:

1. **Listing density** — share of city-wide listings per district
2. **Entire-home ratio** — share of listings that are entire homes
3. **Price level** — median nightly price

```{r tpi}
tpi_data <- listings %>%
  group_by(neighbourhood) %>%
  summarise(
    n_listings   = n(),
    entire_ratio = mean(room_type == "Entire home/apt", na.rm = TRUE),
    median_price = median(price, na.rm = TRUE)
  ) %>%
  filter(n_listings >= 20) %>%
  mutate(
    across(c(n_listings, entire_ratio, median_price),
           ~ (. - min(.)) / (max(.) - min(.)),
           .names = "{.col}_s"),
    TPI = (n_listings_s + entire_ratio_s + median_price_s) / 3
  )

tpi_data %>%
  slice_max(TPI, n = 15) %>%
  mutate(neighbourhood = fct_reorder(neighbourhood, TPI)) %>%
  ggplot(aes(x = TPI, y = neighbourhood, fill = TPI)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_c(option = "A", direction = -1) +
  labs(
    title   = "Tourism Pressure Index - Top 15 Districts",
    subtitle = "Equal-weight: listing density + entire-home ratio + price level",
    x = "TPI (0-1)", y = NULL,
    caption = "Source: Inside Airbnb - Istanbul"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

## Summary

Istanbul's Airbnb market is spatially concentrated, commercially maturing, and recovering strongly after COVID-19. Central tourist districts face the highest tourism pressure, while multi-listing hosts are capturing a growing share of supply.

