---
title: "Data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
```

## About the Dataset

Our primary data source is the **Istanbul Airbnb dataset** from [Inside Airbnb](http://insideairbnb.com/get-the-data). We work with three files:

| File | Approx. Rows | Key Columns |
|------|-------------|-------------|
| `listings.csv` | ~20,000 | id, name, neighbourhood, room_type, price, host_id, reviews |
| `reviews.csv` | ~100,000+ | listing_id, date |
| `calendar.csv.gz` | millions | listing_id, date, available, price |

We also use a **GeoJSON file** of Istanbul neighbourhood boundaries for maps.

### Why This Dataset?

Istanbul attracts over 20 million visitors yearly. Airbnb has grown substantially but its impact on residents is underexplored. This dataset lets us track geographic spread, pricing dynamics, COVID-19 effects, and host commercialisation.

---

## Data Import

```{r load-libraries}
library(tidyverse)
library(janitor)
library(lubridate)
library(scales)
library(sf)
library(knitr)
library(kableExtra)
```

```{r import-data}
listings_raw       <- read_csv("data/listings.csv",    show_col_types = FALSE)
reviews_raw        <- read_csv("data/reviews.csv",     show_col_types = FALSE)
calendar_raw       <- read_csv("data/calendar.csv.gz", n_max = 500000,
                                show_col_types = FALSE)
neighbourhoods_geo <- st_read("data/neighbourhoods.geojson", quiet = TRUE)
```

---

## Data Preprocessing

### Listings

The listings file contains 18 columns. We select the relevant ones and apply these steps:

1. **Price parsing:** Convert `price` to numeric (values are in Turkish Lira, TRY).
2. **Outlier removal:** Drop listings with price below 50 TRY or above 100,000 TRY.
3. **NA handling:** Remove rows missing `price`, `neighbourhood`, or `room_type`.
4. **Date parsing:** Convert `last_review` to date format.

```{r clean-listings}
listings <- listings_raw %>%
  clean_names() %>%
  select(
    id, name, host_id, host_name,
    neighbourhood_group, neighbourhood,
    latitude, longitude,
    room_type, price, minimum_nights,
    number_of_reviews, last_review,
    reviews_per_month,
    calculated_host_listings_count,
    availability_365
  ) %>%
  mutate(
    price       = as.numeric(price),
    last_review = ymd(last_review)
  ) %>%
  filter(
    !is.na(price),
    !is.na(neighbourhood),
    !is.na(room_type),
    price >= 50,
    price <= 100000
  )

cat("Rows after cleaning:", nrow(listings))
```

### Reviews

```{r clean-reviews}
reviews <- reviews_raw %>%
  clean_names() %>%
  mutate(
    date  = ymd(date),
    year  = year(date),
    month = month(date, label = TRUE)
  ) %>%
  filter(!is.na(date))

cat("Rows after cleaning:", nrow(reviews))
```

### Calendar

```{r clean-calendar}
calendar <- calendar_raw %>%
  clean_names() %>%
  mutate(
    date      = ymd(date),
    price     = as.numeric(str_replace_all(price, "[$,]", "")),
    available = available == "t"
  ) %>%
  filter(!is.na(date))

cat("Rows after cleaning:", nrow(calendar))
```

---

## Final Dataset Overview

```{r dataset-summary}
tibble(
  Dataset = c("listings", "reviews", "calendar"),
  Rows    = c(nrow(listings), nrow(reviews), nrow(calendar)),
  Columns = c(ncol(listings), ncol(reviews), ncol(calendar))
) %>%
  kable(caption = "Cleaned dataset dimensions",
        format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r glimpse}
glimpse(listings)
```

---

## Save Final Dataset

```{r save-rdata}
save(listings, reviews, calendar, neighbourhoods_geo,
     file = "data/istanbul_airbnb.RData")
```

Download: [istanbul_airbnb.RData](data/istanbul_airbnb.RData)